---
title: "Precursor_Ion_Scan_XCMS"
output:
  html_document: default
  pdf_document: default
date: "2026-05-27"
---

# Precursor ion scanning of putative strigolactones

In this markdown, we aim at automatizing the detection of peaks from tandem MS data generated using a precursor-ion scan method. This method uses a fixed m/z on the MS2 detector to scan for precursors of a particular fragment. Here we use 97 m/z as bait as it corresponds to the "D-ring" fragment commonly produced by strigolactones.
The data used here is generated by a XEVO TQ XS (Waters) and converted to open source format in MSconvert (ProteoWizard).
Detection of peaks from MS data can be done in different ways. The classical approach uses the centwave algorithm from XCMS using continuous wavelet transform (CWT) or massifquant to detect chromatographic peaks.


## I. Data import and conversion from Waters XEVO TQ XS

### a. Export and convert data

For this analysis, we need:

-Metadata of MS run setup exported from Masslynx to a .csv table
-Experimental design metadata (including root weight if possible)
-Raw mass spec files (.raw from Waters systems) in an open format such as .mzML or .NetCDF

1) Export your metadata by opening your project file in Mass lynx (yourproject.PRO/Data/SampleDB/.SPL) and copy/pasting the method to a blank excel table. Add the headers to the different columns in excel by copy/pasting this: File_Name	Sample_ID	File_Text	MS_File	MS_Tune_File	Inlet_File	Bottle	Inject_Volume	Sample_Type and export to .csv

2) Export the data (yourproject.PRO/Data/*.raw) from the MS computer and convert to open format using MassLynx Databridge (https://support.waters.com/KB_Inf/MassLynx/WKB57100_How_to_export_raw_data_files_to_an_open_file_format_using_a_post_run_process) or MSconvert (https://proteowizard.sourceforge.io/download.html) to a new folder. Don't forget to use the centroiding option on MS convert.

3) Finally, put all these files into a folder of your choice containing a folder with your .raw data and your metadata .csv files.


### b. Load libraries

```{r load libraries, message=FALSE, warning=FALSE}

library(xcms)
library(faahKO)
library(RColorBrewer)
library(pander)
library(magrittr)
library(pheatmap)
library(MSnbase)
library(SummarizedExperiment)
library(ggplot2)
library(foreach)
library(iterators)
library(dplyr)
library(ggpubr)
library(BiocParallel)
library(processx)
library(plotly)
library(htmltools)
library(scales)
library(tidyr)
library(rstatix)
library(purrr)
library(stringr)

```

```{r colors}

cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999")

cbPalette_ordered <- c( "#1f77b4", "#6baed6", "#08306b", "#54278f", "#d73027", "#fdae61", "#f46d43","#FFD700" ,"#3ff000", "#a6d96a", "#66bd63", "#1a9850", "#006837","#808080","#000000", "pink")


```


### c. Import data

Create a new object with the converted data and for the phenodata (metadata) file as follow:

```{r import metadata}

# Set path to metadata directory

PATH <- "/zfs/omics/personal/sjaupit/results/Strigolactone_ID/"

PATH_DATA <- "/zfs/omics/personal/sjaupit/results/Strigolactone_ID/Family_Experiment_PIS/"

# Import metadata

msetup <- read.csv(paste0(PATH,"MassLynx_Setup_FamExp.csv")) #MassLynx setup table

msetup <- msetup %>% dplyr::select(1:2)

meta <- read.csv(paste0(PATH,"Metadata_FamExp.csv")) #Experiment metadata

meta <- merge(msetup, meta, by = "SampleID")



```

```{r import data}

# Get the full path to the .mzML file

msfiles <- dir(path = paste0(PATH_DATA), pattern = ".mzML$",
                  full.names = TRUE)


# Create a phenodata object

pd_temp <- data.frame(File_Name = sub(basename(msfiles), pattern = ".mzML",
                                   replacement = "", fixed = TRUE),
                 stringsAsFactors = FALSE)


pd <- merge(pd_temp, meta, by="File_Name", all.x=TRUE)


```

RUN ONCE

The following step will take time, we save the resulting object in a .rds file to facilitate re-running the code if necessary.

```{r read mzML data, eval = FALSE}


raw_data <- readMSData(files = msfiles, pdata = new("NAnnotatedDataFrame", pd),
                       mode = "onDisk")

#raw_data <- filterRt(raw_data, c(2500, 3500)) #OPTIONAL: set retention time range

head(rtime(raw_data))

saveRDS(raw_data, paste0(PATH,"raw_data_PIS_FamExp.rds")

```


## II. XCMS peak detection

We recommend the following XCMS tutorial if you are not familiar with this package: 

https://jorainer.github.io/metabolomics2018/xcms-preprocessing.html

```{r read object, eval= TRUE}

raw_data <- readRDS(paste0(PATH,"raw_data_PIS_FamExp.rds"))

length(raw_data@processingData@files)

```


### a. OPTIONAL - Pre-processing

```{r smoothing data, optional, eval=FALSE}
# Smooth the signal, then do a simple peak picking.
raw_data_cent <- smooth(raw_data, method = "SavitzkyGolay", halfWindowSize = 2L) %>%
    pickPeaks(refineMz = "descendPeak")

raw_data_cent <- smooth(raw_data, method = "MovingAverage", halfWindowSize = 2)

```

```{r centroiding data, eval=FALSE}

# Define the mz and retention time ranges
serine_mz <- 106.049871
mzr <- c(serine_mz - 0.01, serine_mz + 0.01)
rtr <- c(175, 187)

# Filtering the object
serine <- raw_data_cent %>%
    filterRt(rtr) %>%
    filterMz(mzr)

```

The raw data were imported using the onDisk-mode that does not read the full MS data into memory. Any data manipulation (such as data smoothing or centroiding above) has to be applied on-the-fly to the data each time m/z or intensity values are retrieved. 

To make any data manipulation on an OnDiskMSnExp object persistent we need to export the data to mzML files and re-read the data again. Below we thus save the centroided data as mzML files and import them again.

```{r save centroided files, eval=FALSE}
# Write the centroided data to files with the same names in the current directory

fls_new <- basename(fileNames(raw_data_cent))

writeMSData(raw_data_cent, file = fls_new)

# Read the centroided data

data_cent <- readMSData(fls_new, pdata = new("NAnnotatedDataFrame", pd),
                        mode = "onDisk")

raw_data <- data_cent

```


### b. Initial data inspection

#### MS1

Define colors for the group variable (Species here).

```{r colors for plots}

cbPalette_ordered <- c( "#1f77b4", "#6baed6", "#08306b", "#54278f", "#d73027", "#fdae61", "#f46d43","#FFD700" ,"#3ff000", "#a6d96a", "#66bd63", "#1a9850", "#006837", "#00CD66","#808080","#000000","pink",'#CD00CD', 'green','moccasin',"grey","lightgrey","#8B8878","#E0EEE0","#838B83","#9FB6CD", "#E69F00", "#56B4E9")
                        
order <- c("CuS","CuSh","CuMa","CuM","CuMe","CiLm","CiL","CiLv","LS","CbT","CbF","CbP","CbA","CbM","LC","GP","Control","BLANK","QC","ZM","AT","NB","PS","MT","BO","BD", "P+", "P-")



group_colors <- setNames(
  rep(cbPalette_ordered, length.out = length(order)),
  order
)

```

Choose MS level, here if we target 97 masses in MS2, then we select MS2 level to see only 97 masses.

```{r base peak chromatograms}

bpis1 <- chromatogram(raw_data, aggregationFun = "max", msLevel = 1L) 

```

Plot all chromatograms.

```{r plot base peak chromatograms MS1}

plot(bpis1, 
     col = group_colors[bpis1$Treatment], 
     xaxp  = c(60, 600, 60), 
     yaxp  = c(0, 100000000, 100),
     las = 1, 
     lwd = 1,
     main = "Base Peak Chromatograms (All Samples)")

legend("topleft", legend=unique(bpis1$Treatment), col = group_colors[bpis1$Treatment], cex=0.8, lwd = 1)

```

#### TIC

Here, we look at the TIC per samples and per runs. We can identify failing MS runs and samples.

```{r plot total ion current}

tc <- split(tic(raw_data), f = fromFile(raw_data))
boxplot(tc, col = group_colors[raw_data$Treatment],
        ylab = "intensity", main = "Total ion current", outline = FALSE)

```


### c. Filtering samples

Filter blanks, QC and +P samples.

```{r filter unwanted runs or samples and create new cdfs, eval = TRUE}

filt_pd <- pd[!grepl("BLANK", pd$Treatment),]
filt_pd <- filt_pd[!grepl("QC", filt_pd$Treatment),] 
filt_pd <- filt_pd[!grepl("P[\\+]", filt_pd$Treatment),]
filt_pd <- filt_pd[!grepl("NA", filt_pd$Treatment),]

vector_key <- list(filt_pd$File_Name)
pattern <- unlist(vector_key)

pattern <- paste0(pattern, sep= ".mzML", collapse = "|")

filtcdfs <- list.files(path = paste0(PATH,"Family_Experiment_PIS/"), pattern = pattern, full.names = TRUE, recursive = TRUE)

filt_data <- readMSData(files = filtcdfs, pdata = new("NAnnotatedDataFrame", filt_pd),
                       mode = "onDisk")

```

OPTIONAL

```{r filter retention times, eval = FALSE}

filt_data <- filterRt(filt_data, rt = c(177, 500))

```

```{r remove empty spectra}

filt_data <- filterEmptySpectra(filt_data)

```

```{r base peak chromatograms filtered MS1}

bpis2 <- chromatogram(filt_data, aggregationFun = "max", msLevel = 1L)

```

```{r plot base peak chromatograms filtered MS1}

group_colors2 <- group_colors[!names(group_colors) %in% c("P+","P-")]

plot(
  bpis2,
  col = group_colors2[bpis2$Species],
  xaxp  = c(60, 600, 60), 
  yaxp  = c(0, 100000000, 100),
  las = 1,
  lwd = 1,
  main = "Base Peak Chromatograms (Filtered Samples)"
)
legend(
  "topleft",
  legend = names(group_colors2),
  col = group_colors2,
  cex = 0.4,
  lwd = 1
)

```


### d. Peak detection

#### Calibration peak

This part is used to adjust the peak width and noise parameters. 
Here, we choose a peak from the MS1 chromatogram, it is recommended to test several peaks.

```{r extract one peak}

rtr <- c(360, 380) # Define the rentention time range of the peak area
mzr <- c(200, 500) # Define the m/z range of the peaks

chr_rawFam <- chromatogram(filt_data, mz = mzr, rt = rtr, msLevel = 1L)

```

```{r plot extracted peak}

plot(chr_rawFam, col =  group_colors[chr_rawFam$Species])

```

From the plot above we see that the peak width is between 365 and 380 so we can use 2 as a minimum and 20 as a maximum to allow for margins.
Here we plot the retention time, m/z and intensity of the peak.

```{r plot RT, m/z and intensity, eval=FALSE}

filt_data %>%
    filterRt(rt = rtr) %>%
    filterMz(mz = mzr) %>%
    plot(type = "XIC", layout = NULL)

```


#### Run findChromPeaks function in all chromatograms

This step is the most sensitive, the values of peak width, noise and prefilter must be optimised.
We recommend reading the following vignette: https://rdrr.io/bioc/xcms/man/findChromPeaks-centWave.html

Two peak-picking strategies are tested:

- `CentWaveParam`: classical continuous wavelet transform peak detection,
- `MassifquantParam`: Kalman-filter-based peak detection.

We define the following parameters as follow:

- **Peak width**: expected peak width in seconds.

- **Noise**: value to assign to base noise.

- **Prefilter**: `c(k, I)` with `k` = minimum number of peaks and `I` = intensity.

- **snthresh**: signal-to-noise ratio threshold.

- **integrate**: mexican hat = `1` or real data = `2`.


```{r peak detection in all spectra}

#Set parameters

cwp <- CentWaveParam(peakwidth = c(2, 20), noise = 50000, prefilter = c(1, 1000), snthresh = 7, integrate= 1, fitgauss = TRUE)

mqp <- MassifquantParam(peakwidth = c(2, 20), noise = 20000000,
                     prefilter = c(1, 100000), snthresh = 20, integrate= 1, fitgauss = TRUE)

#Save parameters

saveRDS(cwp, file= paste0(PATH, "Centwave_SLs_FamExp.rds"))

saveRDS(mqp, file=paste0(PATH, "MassifQuant_SLs_FamExp.rds"))

#Run peak

xdataFam_cwp <- findChromPeaks(filt_data, param= cwp) #CentWave peak picking

xdataFam_mqp <- findChromPeaks(filt_data, param= mqp) #massifquant peak picking

#Save peak files

saveRDS(xdataFam_cwp, file=paste0(PATH,"CentWave_Chrompeak_xdataFam.rds"))

saveRDS(xdataFam_mqp, file=paste0(PATH,"MassifQuant_Chrompeak_xdataFam.rds"))

```


## III. Select peak picking algorithm

Here, we select the peak picking algorithm to use and run all chunk below to generate putative peaks tables.

```{r read objects, eval = TRUE}

xdataFam_cwp <- readRDS(file=paste0(PATH,"CentWave_Chrompeak_xdataFam.rds"))

xdataFam_mqp <- readRDS(file=paste0(PATH,"MassifQuant_Chrompeak_xdataFam.rds"))

#Select peaks here (check/uncheck both lines below to choose peak algorithm and set output files names)

xdataFam <- xdataFam_cwp
method <- "Centwave"

#xdataFam <- xdataFam_mqp
#method <- "MassifQuant"


#Add file names extensiong

xdataFam_raw <- xdataFam

head(chromPeaks(xdataFam))


```


### a. Putative peaks pre-processing

Here, we remove failed samples if needed. 
**Be careful to adjust sample numbers to remove between test runs !** 

```{r remove samples without peaks, eval =TRUE}

keep_files <- setdiff(1:length(fileNames(xdataFam)), 44) #Insert file to remove here

xdataFam <- filterFile(xdataFam, file = keep_files, keepAdjustedRtime = keepAdjustedRtime)

filt_pd <- xdataFam@phenoData@data

filt_pd$sample <- seq_len(nrow(filt_pd))

filt_pd <- filt_pd[-44,]

rownames(filt_pd) <- NULL

```


### b. OPTIONAL - Peak refining

```{r merging and refining peaks, eval= FALSE}

mpp <- MergeNeighboringPeaksParam(expandRt = 5, expandMz = 0.1, ppm = 10, minProp = 0.75)

xdataFam <- refineChromPeaks(xdataFam, mpp)

```

The aim of the gap filling step is to reduce the number of such missing values by integrating signals from the original data files for samples in which no chromatographic peak was found from the m/z - retention time region where signal from the ion is expected.

```{r group peaks, eval= FALSE}

pdFamp <- PeakDensityParam(sampleGroups = filt_pd$Species,
                        minFraction = 0.4, bw = 30)

xdataFam <- groupChromPeaks(xdataFam, param = pdFamp)

```

```{r fill gaps, eval = FALSE}

xdataFam <- fillChromPeaks(xdataFam, param = ChromPeakAreaParam())

featureValues(xdataFam, value = "into") |> head()

```

Here we inspect peaks before and after optional pre-processing.

```{r plot reference peak, eval = FALSE}

mzr_1 <- mzr
rtr_1 <- rtr
chr_1 <- chromatogram(filterFile(xdataFam_raw, 14), mz = mzr_1, rt = rtr_1) #No preprocessing
chr_2 <- chromatogram(filterFile(xdataFam, 14), mz = mzr_1, rt = rtr_1) # Pre-processed
par(mfrow = c(1, 2))
plot(chr_1)
plot(chr_2)

```


### c. Print summary

```{r summary}

summary_fun <- function(z)
    c(peak_count = nrow(z), rt = quantile(z[, "rtmax"] - z[, "rtmin"]))

sumX <- lapply(split.data.frame(
    chromPeaks(xdataFam), f = chromPeaks(xdataFam)[, "sample"]),
    FUN = summary_fun)

sumX <- do.call(rbind, sumX)

sumX <- as.data.frame(sumX)

sumX$sample <- as.numeric(rownames(sumX))

sumX_annot <- merge(
    sumX,
    filt_pd[, c("sample", "SampleID")],
    by = "sample",
    all.x = TRUE
)

head(sumX_annot)

```


## IV. Identification of putative SLs peaks

#### a. Inspection of indentified peaks


```{r create peak table, eval = TRUE}

xdataFam_pp <- xdataFam

putative_peaks_Fam_std <- as.data.frame(chromPeaks(xdataFam_pp))

head(chromPeaks(xdataFam_pp))

```

Here we plot the chromatograms and signal to noise ratios of the peaks.

```{r per file}

plotChromPeaks(xdataFam_pp, file = 10)

```

```{r merge tables peaks}

putative_peaks_Fam_std$sample <- as.character(putative_peaks_Fam_std$sample)
putative_peaks_Fam_std <- merge(filt_pd, putative_peaks_Fam_std , by = "sample")

```


### b. Peak filtering

Here we calculate the baseline threshold for the noise.

```{r peak intensity distribution and baseline threshold}


x <- putative_peaks_Fam_std$into

h <- hist(x, breaks = 1000, plot = FALSE)

smooth_counts <- smooth.spline(h$mids, h$counts, spar = 0.7)

y <- predict(smooth_counts, h$mids)$y

peak_idx <- which.max(y)

dy <- diff(y)

min_idx <- which(diff(sign(dy)) == 2)
min_idx <- min_idx[min_idx > peak_idx][1]

baseline_threshold <- h$mids[min_idx]

baseline_threshold <- 0.75*baseline_threshold

print(baseline_threshold)

plot(h$mids, h$counts, type = "l", main = "Baseline threshold detection")
lines(h$mids, y, col = "blue", lwd = 2)
abline(v = baseline_threshold, col = "red", lwd = 2)




```

Here we filter the peaks table to identify putative SLs peaks. This step is crucial to obtain meaningful peaks and different combination of parameters should be tested.

```{r filter peaks}

putative_peaks_Fam_std$rmz <- round(putative_peaks_Fam_std$mz, digit=1)

putative_peaks_Fam_std$rrt <- (round(putative_peaks_Fam_std$rt, digit=0))/60

#putative_peaks_Fam_std_filt <- putative_peaks_Fam_std %>% filter(sn > 10) #signal/noise

putative_peaks_Fam_std_filt <- putative_peaks_Fam_std %>% filter(into > baseline_threshold)

#putative_peaks_Fam_std_filt <- putative_peaks_Fam_std_filt %>% filter(rmz < 420)
#putative_peaks_Fam_std_filt <- putative_peaks_Fam_std_filt %>% filter(rmz > 330)


putative_peaks_Fam_std_filt <- putative_peaks_Fam_std_filt %>% filter(Species != "CuSh")
#putative_peaks_Fam_std_filt <- putative_peaks_Fam_std_filt %>% filter(Species != "CuM")
#putative_peaks_Fam_std_filt <- putative_peaks_Fam_std_filt %>% filter(Species != "CuMa")



#putative_peaks_Fam_std_filt <-  putative_peaks_Fam_std_filt %>% group_by(rmz, rrt) %>% filter(n()>1) #remove low occurrence peaks

guests <- as.vector(c("BD", "AT", "BO", "PS", "MT","NB", "SL", "ZM"))

putative_peaks_Fam_std_filt <- putative_peaks_Fam_std_filt[!putative_peaks_Fam_std_filt$Species %in% guests,]

```

### c. Plot peaks

Plot filtered peaks pre-filtering.

```{r plot raw peaks, eval = TRUE}

plot1 <- ggplot(putative_peaks_Fam_std, aes(x=rt, y=rmz, color=Species, size = into))+geom_count()+labs( y = "m/z", title ="MS results")  + theme(axis.text.x = element_text(angle = 90)) + theme(text = element_text(size = 19))+ 
  geom_text(aes(label = rmz), color= "#000000" , nudge_y = -5, check_overlap = TRUE, size = 5)  +
scale_x_continuous(breaks = scales::pretty_breaks(n = 20))

plot1

```

Here we normalize peak intensities by root weight before plotting.

```{r root weight normalize}

putative_peaks_Fam_std_filt_combined <- as.data.frame(putative_peaks_Fam_std_filt)


putative_peaks_Fam_std_filt_combined$norm <- putative_peaks_Fam_std_filt_combined[,"into"]/putative_peaks_Fam_std_filt_combined[,"Weight"]

names(putative_peaks_Fam_std_filt_combined)[names(putative_peaks_Fam_std_filt_combined) == 'into'] <- 'Intensity'


```

Plot putative peaks post-filtering.

```{r plot peaks filtered}

cbPalette_ordered <- c( "#1f77b4", "#6baed6", "#08306b", "#54278f", "#d73027", "#fdae61", "#f46d43","#FFD700" ,"#3ff000", "#a6d96a", "#66bd63", "#1a9850", "#006837","#808080","#000000", "pink")


     
order <- c(
  "Cucumis_sativus",
  "Cucumis_sativus_var._hardwickii",
  "Cucumis_melo_subsp._agrestis",
  "Cucumis_melo",
  "Cucumis_metulifer",
  "Citrullus_mucosospermus",
  "Citrullus_lanatus",
  "Citrullus_lanatus_subsp._vulgaris",
  "Lagenaria_siceraria",
  "Cucurbita_pepo_subsp._texana",
  "Cucurbita_pepo_subsp._fraterna",
  "Cucurbita_pepo_var._melopepo",
  "Cucurbita_argyrosperma",
  "Cucurbita_maxima",
  "Luffa_aegyptiaca",
  "Gynostemma_pentaphyllum",
  "Control"
)

putative_peaks_Fam_std_filt_combined$Latin <- factor(putative_peaks_Fam_std_filt_combined$Latin, levels = order)

plot2 <- ggplot(putative_peaks_Fam_std_filt_combined, aes(x=rrt, y=rmz, color=Latin, size = Intensity))+geom_count()+labs( y = "m/z", title ="MS results", color = "Species")  + theme(axis.text.x = element_text(angle = 90)) + theme(text = element_text(size = 19))+ scale_color_manual(values = cbPalette_ordered) +
  geom_text(aes(label = rmz), color= "#000000" , nudge_y = -5, check_overlap = TRUE, size = 5) + xlab("Retention time (min)") 


plot2

ggsave(plot2, file = paste0(PATH,"Putative_Peaks_Cucurb_",method,".png"), width = 12, height = 7)

```


```{r 3D plots}

figFamExfilt <- plot_ly(putative_peaks_Fam_std_filt_combined, x = ~rrt, y = ~rmz, z = ~Intensity, color = ~Latin, colors = cbPalette_ordered) %>%
  config(
    edits = list(
      annotationPosition = TRUE,
      annotationTail = TRUE,
      annotationText = TRUE
    ),
    toImageButtonOptions = list(format = "svg")
  )

figFamExfilt


```

### d. Peak binning and control filtering

Here, we bin peaks by retention times and m/z and highlight peaks found in control pots without plants.

```{r bin peaks and filter controls}

#Set parameters

rt_bin_size  <- 0.16  # retention time window
mz_bin_size  <- 2     # m/z window

control_name <- "C"   # control label
min_presence <- 0.05   # minimum fraction of replicates containing signal
control_max  <- 0.05  # maximum fraction allowed in controls

# Prepare data

dat <- putative_peaks_Fam_std_filt_combined %>%
  mutate(
    rmz = as.numeric(rmz),
    rrt = as.numeric(rrt),

    # RT bins
    rt_bin = floor(rrt / rt_bin_size) * rt_bin_size + (rt_bin_size),

    # MZ bins
    mz_bin = floor(rmz / mz_bin_size) * mz_bin_size + (mz_bin_size/2),

    feature_bin = paste0(
      "RT_", round(rt_bin, 2),
      "_MZ_", round(mz_bin, 1)
    )
  )


feature_summary <- dat %>%
  group_by(
    SampleID,
    Species,
    feature_bin,
    Latin,
    rt_bin,
    mz_bin
  ) %>%
  summarise(
    intensity = mean(maxo, na.rm = TRUE),
    detected  = any(maxo > 0, na.rm = TRUE),
    .groups = "drop"
  )


# Remove features in control samples

feature_presence <- feature_summary %>%
  group_by(feature_bin, Species) %>%
  summarise(
    n_samples = n(),
    prevalence = mean(detected),
    mean_intensity = mean(intensity, na.rm = TRUE),
    .groups = "drop"
  )

# Presence in controls

control_features <- feature_presence %>%
  dplyr::filter(Species == control_name) %>%
  dplyr::select(feature_bin, control_prev = prevalence)

control_features

# Presence in plants

treated_features <- feature_presence %>%
  filter(Species != control_name)

# Keep features that are absent in controls

selected_features <- treated_features %>%
  left_join(control_features, by = "feature_bin") %>%
  mutate(control_prev = replace_na(control_prev, 0)) %>%
  dplyr::filter(
    prevalence >= min_presence,
    control_prev <= control_max
  ) %>%
  pull(feature_bin) %>%
  unique()

feature_filtered <- feature_summary %>%
  filter(feature_bin %in% selected_features)

feature_filtered <- as.data.frame(feature_filtered)

feature_filtered <- feature_filtered[
    order(feature_filtered$rt_bin, decreasing = FALSE),
]

# Unique mass transitions and RT

unique_features <- unique(
  feature_filtered[, c("feature_bin", "rt_bin", "mz_bin")]
)

rownames(unique_features) <- NULL

print(unique_features)

# Plot intensity of putative features

plot_data <- feature_filtered

plotArea <- ggplot(
  plot_data,
  aes(
    x = feature_bin,
    y = intensity
  )
) +
  geom_boxplot(outlier.size = 0.4) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 0.7) +
  labs(
    title = "Differential RT/MZ Features",
    x = "",
    y = "Peak Intensity"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 8)
  )

plotArea

# Summary tables

feature_presence_table <- plot_data %>%
  group_by(
    feature_bin,
    Latin,
    Species
  ) %>%
  summarise(
    n_samples = n_distinct(SampleID),
    mean_intensity = mean(intensity, na.rm = TRUE),
    max_intensity = max(intensity, na.rm = TRUE),
    detected = any(intensity > 0, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(detected)

# Print long-format table
print(feature_presence_table)

feature_presence_table <- as.data.frame(feature_presence_table)


feature_presence_wide <- feature_presence_table %>%
  mutate(present = "YES") %>%
  unite(group, Latin,  sep = " | ") %>%
  dplyr::select(feature_bin, group, present) %>%
  pivot_wider(
    names_from = group,
    values_from = present,
    values_fill = "-"
  )

print(feature_presence_wide)

feature_presence_wide <- as.data.frame(feature_presence_wide)

```


### e. Save outputs

```{r save tables and graphs, eval = TRUE}

htmlwidgets::saveWidget(as_widget(figFamExfilt), paste0(PATH,"figFamExfilt_",method,".html"))

write.csv(putative_peaks_Fam_std, paste0(PATH,"putative_peaks_Fam_std_all_samples_",method,".csv"))

write.csv(putative_peaks_Fam_std_filt, paste0(PATH,"putative_peaks_Fam_std_filt_all_samples_",method,".csv"))

write.csv(putative_peaks_Fam_std_filt_combined, paste0(PATH,"putative_peaks_Fam_std_filt_combined_all_samples_",method,".csv"))

write.csv(feature_presence_table, paste0(PATH,"feature_presence_table_",method,".csv"))

write.csv(feature_presence_wide, paste0(PATH,"feature_presence_wide_",method,".csv"))

write.csv(feature_filtered, paste0(PATH,"putative_features_per_plants_",method,".csv"))

write.csv(unique_features, paste0(PATH,"putative_features_info_",method,".csv"))

```


### f. Compare methods

Here we compare features found by both peak picking algorithms.

```{r compare features per algorithms, eval = TRUE}

cwp_features <- read.csv(paste0(PATH,"putative_features_info_Centwave.csv"))

mqp_features <- read.csv(paste0(PATH,"putative_features_info_MassifQuant.csv"))

cwp_bin <- cwp_features$feature_bin
mqp_bin <- mqp_features$feature_bin

common_features <- intersect(cwp_bin, mqp_bin)

length(common_features)

common_mz <- sub(".*MZ_", "", common_features)

unique(common_mz)


only_cwp <- setdiff(cwp_bin, mqp_bin)
only_mqp <- setdiff(mqp_bin, cwp_bin)


comparison <- list(
  common = common_features,
  only_cwp = only_cwp,
  only_mqp = only_mqp
)



cat("\n================ FEATURE COMPARISON SUMMARY ================\n\n")

cat("- Common features (shared between methods):", length(common_features), "\n")
cat("- Unique to CentWave:", length(only_cwp), "\n")
cat("- Unique to MassifQuant:", length(only_mqp), "\n\n")

cat("- Unique m/z transitions in common features:", unique(common_mz), "\n\n")

cat("============================================================\n")

```

